DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-03-14 R/processing-code/MADAproject_pt2_ruiz.qmd
Pt 3
updated qmd and manuscript

## This file contains the process for loading, cleaning and exploring the datasets.
# Loading libraries
# This list contains the specific versions used in the script.
```{r, message=FALSE, warning=FALSE}
library(readxl) # for loading Excel files (version 1.4.0)
library(httr) # for getting data from the web (version 1.4.2)
library(dplyr) # for data processing/cleaning (version 1.0.7)
library(tidyr) # for data processing/cleaning (version 1.1.3)
library(skimr) # for visualizing data (version 2.1.3)
library(here) # allows relative pathways for data (version 1.0.1)
library(readr) # reading and writing data to and from CSV, TSV, and fixed-width files (version 2.0.1)
library(naniar) # tools to work with missing data (version 0.6.1)
library(ggplot2) # plot creation (version 3.3.5)
library(sf) # standardized way to encode spatial data (version 1.0-2)
library(topicmodels) # tools for topic modeling on text data (version 0.2-12)
library(openxlsx) # Facilitates reading from and writing to .xlsx files (version 4.2.4)
library(lubridate) # Simplifies the work with dates and times (version 1.7.10)
library(knitr) # Allows for dynamic report generation in R (version 1.33)
library(kableExtra) # generates simple HTML or LaTeX tables (version 1.3.4)
library(webshot2) # tools to take screenshots of web pages (version 0.0.2)
library(magick) # advanced image processing in R (version 2.7.3)
library(leaflet) # Enables the creation of interactive maps (version 2.0.4.1)
library(htmlwidgets) # Facilitates the creation of R bindings to JavaScript libraries (version 1.5.4)
library(suncalc) # Calculates sunset and sunrise times based on location (version 0.5.1)
```

@@

#load the first few rows of data to verify the data loaded as expected
head(trap_location)
str(trap_location)

#The last 3 columns have missing data. However, they are not important for the purposes of this project
# there is not a need to delete the records
#However, we will delete the columns

trap_location_selected <- trap_location %>%
  select(`Trap Site Address`, MCD, Location, DecLat, DecLong)
trap_location_selected

@@

## Mosquito virus data

### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020

### This code will load all the files, combine them, and then save the combined file in the "raw data" folder

```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

@@

## Weather data

### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023

```{r}
# Construct the path to the file
file_path_wx = here("data", "processed-data", "weather_airport.csv")

@@

## Human and veterinary cases of eastern equine encephalitis and West Nile virus

```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "processed-data", "human_cases.xlsx")

@@

#### The mosquito dataset spans 2007 to 2023. However, the virus and case data only covers the years 2014 to 2020. Because mosquito populations can be influences by processes of the previous year. We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.

### Identify missing data

```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records.The other variable are important and so we will remove those records with the other missing data.

### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with accepted standards, I will change them back to *Aedes*.

```{r}
# Get unique values using dplyr
unique_genus_values <- mosquito_raw %>%
  distinct(genus) %>%
  pull(genus)

@@

# Check the first few rows to confirm changes, or use unique() to see the change in genus values
unique(mosquito_raw$genus)

### Execute proposed changes.

```{r}
# Create a copy of the dataframe to avoid modifying the original
mosquito_modified <- mosquito_raw

# Change records where genus is "Ochlerotatus" to "Aedes"
#mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"

# Filter out records where Collection Date is before 2013 or after 2021
# Ensure Collection Date is in Date format if it's not already
mosquito_modified$`collection date` <- as.Date(mosquito_modified$`collection date`)

# Convert all species code values to uppercase
mosquito_modified <- mosquito_modified %>%
  mutate(`species code` = toupper(`species code`))

# Now apply the date filter
mosquito_modified <- mosquito_modified[mosquito_modified$`collection date` >= as.Date("2013-01-01") & 
                                       mosquito_modified$`collection date` <= as.Date("2021-12-31"), ]

# Remove rows with NA values in the remaining columns of the modified dataframe, 
# except for "Submitted for Testing" which we'll handle separately
temp_columns <- names(mosquito_modified)
temp_columns <- temp_columns[temp_columns != "submitted for testing"]

# Applying NA omit on selected columns only
mosquito_clean_except_submitted <- na.omit(mosquito_modified[temp_columns])

# Since na.omit on selected columns messes with row alignment, let's filter the original dataset based on the cleaned IDs
clean_ids <- row.names(mosquito_clean_except_submitted)
mosquito_final <- mosquito_modified[clean_ids, ]

# Add back "Submitted for Testing" from the original dataframe if needed
# Note: This step might be redundant if "Submitted for Testing" was not actually removed or if it's included in the cleaning process
mosquito_final$`submitted for testing` <- mosquito_raw$`submitted for testing`[clean_ids]

@@

## Reformat the collection data

### The Massachusetts Department of Public Health requires that all mosquito control district submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.

### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.

```{r}
library(dplyr)
library(tidyr)

# Group and summarize the mosquito data to prepare for wide format transformation.
# This step aggregates the total 'pool size' for each unique combination of 
# 'collection date', 'town', 'trap type', 'species code', and 'mcd' (Mosquito Control District),
# while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
mosquito_summarized <- mosquito_final %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
   # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.

@@

# Exploring the Data

##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.

```{r}
# Create the map
map_study <- leaflet(trap_location) %>%
  addTiles() %>%  # This adds the default OpenStreetMap tiles
  addMarkers(lng = ~DecLong, lat = ~DecLat, popup = ~Location)  # Customize the popup content as needed

@@

## EEE in mosquitoes and mammals

### To give an idea of when EEE virus silled over from mosquitoes and birds and into mammal populations, lets take a look at the data. Let's see when the first EEE isolations were found in mosquitoes each year. Let's also look at the years that human and veterinary cases occured.

### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020. 2019 had the highest case counts for both humans and other mammals. **Because of this, we will pay close attention to 2019 through out the exploration.**

```{r}
library(dplyr)
library(lubridate)

@@

## Which county has the most EEE positive mosquitoes of those tested?

```{r}
eee_positive_by_county <- combined_virus_data %>%
  filter(Virus == "EEE") %>%  # Filter for rows where Virus is EEE
  group_by(County) %>%
  summarise(EEE_Positive_Tests = n()) %>%  # Count the number of EEE positive tests in each group
  arrange(desc(EEE_Positive_Tests))  # Order the results by EEE_Positive_Tests in descending order

@@

## *Culiseta melanura* (MEL) is the main enzootic vector of EEE. Some entomologist believe that high levels of MEL traps indicate higher risk for EEE transmission. Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.

```{r}
library(dplyr)
library(ggplot2)
library(readr) # For read_csv
library(here)

@@

## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame.

```{r annual precip}
# Convert DATE from character to Date type using as.Date and specifying the format
wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")

>>>>

## This file contains the process for loading, cleaning and exploring the datasets.
# Loading libraries
# This list contains the specific versions used in the script.
```{r, message=FALSE, warning=FALSE}
library(readxl) # for loading Excel files (version 1.4.0)
library(httr) # for getting data from the web (version 1.4.2)
library(dplyr) # for data processing/cleaning (version 1.0.7)
library(tidyr) # for data processing/cleaning (version 1.1.3)
library(skimr) # for visualizing data (version 2.1.3)
library(here) # allows relative pathways for data (version 1.0.1)
library(readr) # reading and writing data to and from CSV, TSV, and fixed-width files (version 2.0.1)
library(naniar) # tools to work with missing data (version 0.6.1)
library(ggplot2) # plot creation (version 3.3.5)
library(sf) # standardized way to encode spatial data (version 1.0-2)
library(topicmodels) # tools for topic modeling on text data (version 0.2-12)
library(openxlsx) # Facilitates reading from and writing to .xlsx files (version 4.2.4)
library(lubridate) # Simplifies the work with dates and times (version 1.7.10)
library(knitr) # Allows for dynamic report generation in R (version 1.33)
library(kableExtra) # generates simple HTML or LaTeX tables (version 1.3.4)
library(webshot2) # tools to take screenshots of web pages (version 0.0.2)
library(magick) # advanced image processing in R (version 2.7.3)
library(leaflet) # Enables the creation of interactive maps (version 2.0.4.1)
library(htmlwidgets) # Facilitates the creation of R bindings to JavaScript libraries (version 1.5.4)
library(suncalc) # Calculates sunset and sunrise times based on location (version 0.5.1)
library(stringr) # Makes string manipulation easier and more consistent (version 1.5.1)
library(ISOweek) # Calculates week number according to ISO 8601 standards (version 0.6-2)
```

@@

#load the first few rows of data to verify the data loaded as expected
head(trap_location)

#The last 3 columns have missing data. However, they are not important for the purposes of this project
# there is not a need to delete the records
# this is a free text field that provides notes about the location.
# they are not needed so we will delete them
trap_location_selected <- trap_location %>%
  select(`Trap Site Address`, MCD, Location, DecLat, DecLong)
trap_location_selected

@@

## Mosquito virus data
### The data includes all the positive PCR results for EEE and WNV for all mosquitoes submitted for testing
### These files were obtained from: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
### The virus isolation data is stored in separate spreadsheets by year from 2014 to 2020
### This code will load all the files, combine them, and then save the combined file in the "raw data" folder
```{r}
#Define the path to the folder with Excel files
folder_path <- here("data", "processed-data", "virus_data")

@@

## Weather data
### temperature and precipitation are important factors in mosquito population dynamics
### This dataset contains daily weather summaries from the Taunton, MA Airport from 2007-2023
```{r}
# Construct the path to the file
file_path_wx = here("data", "processed-data", "weather_airport.csv")

@@

## Human and veterinary cases of eastern equine encephalitis and West Nile virus
### these were extracted from annual mosquito reports publised here: https://www.mass.gov/lists/arbovirus-surveillance-plan-and-historical-data
### human cases are listed by county only. animal cases are listed by town/city (with the exception of 1 year)
```{r}
# Construct the path to the human case data file
file_path_human <- here("data", "processed-data", "human_cases.xlsx")

@@

#### The mosquito dataset spans 2007 to 2023. However, the virus and case data only covers the years 2014 to 2020. Because mosquito populations can be influenced by processes of the previous year. We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.

### Identify missing data
```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

#### The majority of the missing data is in the "Submitted for Testing" variable. This is not a crucial variable, so we can keep those records.The other variable are important and so we will remove those records with the other missing data.
### The Massachussetts Department of Public Health adpoted a taxonomy change before it was fully vetted. The change divided the genus *Aedes* into a new genus: *Ochlerotatus*. To align with accepted standards, I will change them back to *Aedes*.
```{r}
# Get unique values using dplyr
unique_genus_values <- mosquito_raw %>%
  distinct(genus) %>%
  pull(genus)

@@

# Check the first few rows to confirm changes, or use unique() to see the change in genus values
unique(mosquito_raw$genus)
```

### Execute proposed changes.
```{r}
# Create a copy of the dataframe to avoid modifying the original
mosquito_modified <- mosquito_raw

# Change records where genus is "Ochlerotatus" to "Aedes" to adhere to current taxonomy standards
mosquito_modified$genus[mosquito_modified$genus == "Ochlerotatus"] <- "Aedes"

# Ensure Collection Date is in Date format 
mosquito_modified$`collection date` <- as.Date(mosquito_modified$`collection date`)

# Convert all species code values to uppercase
# This will minimize issues were data as entered incorrectly
# for example Mel, mel, and MEL
mosquito_modified <- mosquito_modified %>%
  mutate(`species code` = toupper(`species code`))

# Remove rows with NA values in the remaining columns of the modified dataframe, 
# except for "Submitted for Testing" which we'll handle separately
# when Submitted for testing is NA or missing, it indicates that there were no mosquitoes captured during a trap event.
# This is common at the beginning and end of season and also when a trap malfunctions
temp_columns <- names(mosquito_modified)
temp_columns <- temp_columns[temp_columns != "submitted for testing"]
mosquito_modified <- na.omit(mosquito_modified, cols = temp_columns)

# Add "week_num" column
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
mosquito_modified <- mosquito_modified %>%
  mutate(
    # Directly format the collection date into "YYYY-WW"
    week_num = paste0(year(`collection date`), "-", 
                      sprintf("%02d", isoweek(`collection date`)))
  )

# before filtering the dates to have a cleaned dataset with all the records.
write_csv(mosquito_modified, here("data", "processed-data", "mosquito_clean_all.csv"))
saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))

# Now apply the date filter to match the analytical horizon of the project
mosquito_final <- mosquito_modified %>%
  filter(`collection date` >= as.Date("2013-01-01") & `collection date` <= as.Date("2021-12-31"))

@@

## Reformat the collection data
### The Massachusetts Department of Public Health and US CDC require that all mosquito control districts submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
### The resulting data frame will rectangle in shape and contain 1 row per trap event with all possible species as a separate column. 
```{r}
# Group and summarize the mosquito data to prepare for wide format transformation.
# This step aggregates the total 'pool size' for each unique combination of 
# 'collection date', 'town', 'trap type', 'species code', and 'mcd' (Mosquito Control District),
# while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
mosquito_summarized <- mosquito_final %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
   # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.

@@

# Exploring the Data
##Study location Below is a map of the mosquito trap sites. This is an interactive map and will only work on web or other html-friendly platforms.
```{r}
# Create the map
map_study <- leaflet(trap_location) %>%
  addTiles() %>%  # This adds the default OpenStreetMap tiles
  addMarkers(lng = ~DecLong, lat = ~DecLat, popup = ~Location)  # Customize the popup content as needed

@@

## EEE in mosquitoes and mammals
### To give an idea of when EEE virus spilled over from mosquitoes and birds and into mammal populations, lets take a look at the data. Let's see when the first EEE isolations were found in mosquitoes each year. Let's also look at the years that human and veterinary cases occured.
### From the table, we see that human or othe mammal cases were recorded in 2014, 2018, 2019, and 2020. 2019 had the highest case counts for both humans and other mammals. **Because of this, we will pay close attention to 2019 through out the exploration.**
```{r}
library(dplyr)
library(lubridate)

@@

## Which county has the most EEE positive mosquitoes of those tested?
```{r}
eee_positive_by_county <- combined_virus_data %>%
  filter(Virus == "EEE") %>%  # Filter for rows where Virus is EEE
  group_by(County) %>%
  summarise(EEE_Positive_Tests = n()) %>%  # Count the number of EEE positive tests in each group
  arrange(desc(EEE_Positive_Tests))  # Order the results by EEE_Positive_Tests in descending order

@@

## *Culiseta melanura* (MEL) is the main enzootic vector of EEE. Some entomologist believe that high levels of MEL traps indicate higher risk for EEE transmission. Using the new mosquito collection table created from the pivot, let's calculate the proportion of MEL captured compared to all other mosquito vector species and create a plot to visualize it.
```{r}
library(dplyr)
library(ggplot2)
library(readr) # For read_csv
library(here)

@@

## Mosquito abundance is affected by weather. Mosquitoes depend on standing water to complete their larval development. Let's take a look at the annual precipitation using the wx data frame.
```{r annual precip}
# Convert DATE from character to Date type using as.Date and specifying the format
wx$DATE <- as.Date(wx$DATE, format="%m/%d/%y")

### 2024-03-14 products/manuscript/Manuscript.qmd
Pt 3
updated qmd and manuscript

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for Culiseta melanura
(MEL), which is recognized as the primary enzootic vector for Eastern
Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
that hold water throughout the year. These crypts serve as a main
habitat for MEL larval growth.

@@

The cornerstone of this project is mosquito surveillance data. This
dataset includes mosquito counts by species for all mosquito traps set
in Bristol County, MA between 2007 to 2023. In addition to mosquito
counts by species, the dataset includes the town or city in which they
were collected, the collection data, and whether the mosquitoes were
submitted for virus testing.

@@

\*This project will lay the foundation for my dissertation research.
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.**

>>>>

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for *Culiseta melanura*
(MEL), which is recognized as the primary enzootic vector for Eastern
Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
that hold water throughout the year. These crypts serve as a main
habitat for MEL larval growth.

@@

The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The BCMCP, perform mosquito surveillance from June through October annually. During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites each week. Some of the sentinel site have been used for over 50 years. The traps are installed in the afternoon and retrieved the following morning. Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance.

Additionally, the county employs gravid traps, which are specifically aimed at attracting and capturing gravid *Culex* mosquitoes looking for egg-laying sites. While these traps are primarily used for West Nile Virus (WNV) surveillance, they hold significant value for EEE monitoring as well. Given the role of *Culex* species as bridge vectors for EEE, their capture in gravid traps provides critical insight into the risk of virus transmission.

The  In addition to mosquito
counts by species, the dataset includes the town or city in which they
were collected, the collection data, and whether the mosquitoes were
submitted for virus testing.

@@

\*This project will lay the foundation for my dissertation research.
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** From prior
research we know that the mosquito life cyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. Precipitation increases the
amount of standing water and increases the chances of larval survival
during the mosquito season and also during the winter months for those
species that overwinter in the larval stage. For EEE, the main enzootic
vector, MEL, overwinters as larvae.

While these relationships are well known, they are not well described
for EEE, especially in terms of quantifying the role these two factors
play in EEE transmission. Specifically, there's a gap in our
understanding of how these factors influence the number of mosquitoes
caught in traps and the rate of virus infection among those mosquitoes.
This gap is more pronounced when distinguishing the impacts on enzootic
(mosquitoes that primarily feed on birds) versus bridge vector
(mosquitoes that can transmit the virus from birds to humans) mosquito
species. Even less well described is the effect the factors have on
enzootic vs. bridge vector mosquito species.

Moreover, the influence of other variables, such as bird populations,
remains even less explored. Birds, particularly species of passerines,
serve as amplifying hosts for EEEV and WNV, potentially affecting the
level of virus present in an area. Despite their importance in the virus
transmission cycle, the effect of fluctuations in bird populations on
the prevalence of these viruses throughout a season has received minimal
attention. Finally, given the migratory nature of many passerine
species, it's worth exploring whether EEEV activity in the warmer
regions of the United States, during the months preceding mosquito
season in Massachusetts, could serve as an early indicator of virus
activity in the state. Understanding these dynamics is crucial for
predicting outbreaks and implementing effective control measures.
